---
title: 'Scale'
date: "14 May 2022 (first version 28 Jun 2021)"
output: 
  html_document:
    code_folding: hide
bibliography: bibliography_scale.bib
---


```{r setup, include=TRUE, message=FALSE, warning=FALSE} 

#Preliminaries:

knitr::opts_chunk$set( message=FALSE, warning=FALSE) #echo = FALSE,

rm(list=ls())

library(tidyverse)
library(viridis)
library(scales)
library(stargazer)
library(WDI)
library(plotly)


theme_set(theme_light())

dir.create("filesScale", showWarnings = FALSE)


```

## Organisms & Organizations

Size matters -- in many situations the formula $Y=aX^\beta$ can describe the empirical association between observed outcomes (Y) and size (X), e.g. for living organisms life expectancy (Y) is found to scale as size (X) to the power of 1/4. Geoffrey West argues in his extraordinary book 'Scale' [@west2017scale] that networks are a key aspect in such power laws in the biological domain, but also in cities and companies. Crucial for his analysis is that networks (such as vessels in animals, or pipelines in cities) are 1. invariant at their end points, 2. space filling and 3., optimization is taking place (evolution). For the biological domain nonlinear scaling is derived for a vast number of attributes based on these 3 assumptions, eg. there is a nonlinear association between between animal size and blood pressure. Furthermore, according to West the literature seems to suggest that, after taking the logarithms of variables, these correlations regularly involve number 4. West's discussion of the role of scaling laws in the biological domain suggests that the magic number is rather 4 than 42, as suggested by @adams2017ultimate. ;)

## Scaling Laws in Practice

Lets have a look at data from the real world [@allison1976sleep]. According to West the double-log association between life span and body size should be about 1/4. When visualizing the correlation we apply the common logarithm (base 10) to both variables.

```{r }
#sleep.txt generated from http://lib.stat.cmu.edu/datasets/sleep
#Data import and manipulation:
sleep_raw <- read_lines(
  "filesScale/sleep.txt",
  skip = 50,
  n_max = 62
)

mplt_slr <- str_locate_all(sleep_raw, "\\s[\\d\\.\\-]") 

sleep_index <- c(1:length(sleep_raw))  

for (i in sleep_index) {
  for (j in c(1:10)) {
  str_sub(sleep_raw[i],mplt_slr[[i]][j,1], mplt_slr[[i]][j,1]) <- ","
  
}}

sleep_df <- as.data.frame(sleep_raw) %>%
          separate( col=1, sep=",",  into=c("species", "body_weight_kg", "brain_weight_g", "hrs_slow_wave_sleep", "hrs_paradoxical_sleep", "hrs_total_sleep", "maximum_life_span", "gestation_time", "predation_index", "sleeping_exp._index", "overall_danger_index")) %>%
          mutate(body_weight_kg=as.numeric(body_weight_kg), maximum_life_span=as.numeric(maximum_life_span)) %>% 
          filter(body_weight_kg!=-999 & maximum_life_span!=-999)        

#How does data look like?
sleep_df %>%
  select(species, maximum_life_span, body_weight_kg) %>%
  top_n(3) %>%
  stargazer(summary=FALSE, type="text", title="Table 1: Body weight data")

#plot scaled association between life span and life expectancy:
sleep_df %>%
  ggplot(aes(y=maximum_life_span, x=body_weight_kg)) + #, color=species
  geom_point() +
  geom_smooth(method=lm, se=FALSE, aes(color = NULL)) +
  scale_x_log10(labels = comma) + #
  scale_y_log10() +
  scale_colour_viridis_d(option = "turbo") +
  theme_bw() +
  labs(title="Figure 1: Body weight and life-span", y="log(life span in years)", x="log(body weight in kg)")

```

There is a lot of variability in the data, the slope is 0.211. But we see an exponential association of 1/4 cannot be rejected statistically with the data at hand:

```{r }
lm(log(maximum_life_span) ~ log(body_weight_kg), data=sleep_df) %>%
  stargazer(type="text", title="Table 2: Association between life-span & body weight", align=TRUE, ci=TRUE)

```

According to West, similar to animals, cities and even companies can be described by their network structure (people are final leaves in a network). In the context cities, there is sublinear scaling (0.85) for the association between population and infrastructure and superlinear scaling (1.15) in the association between population and e.g. wages. However, other research suggests that such findings for cities may not solely be due to optimization by intra-city processes [@ribeiro2021association]. 

Power laws are extremely relevant in many aspects of life, and they can emerge due to different mechanisms [@andriani2006relevance], but underlying mechanisms are not always known yet. 
The role of power laws in the success of companies, for example, can be explained with matching: a large and successful company can hire the best employees which leads to even higher economic outcomes [@gabaix2016power]. Power laws even emerge on the country level as shown in the following. Let us use Worldbank data from 2019 to reveal the empirical association between countries' income (GDP) and population size:  

```{r fig.asp = 0.8, fig.width = 8.5} 

# #read WB-Data from 2019:
# wdi_in <- WDI(
#   country = "all",
#   indicator = c("NY.GDP.MKTP.CD", "SP.POP.TOTL", "SP.URB.TOTL.IN.ZS"),
#   start = 2019,
#   end = 2019,
#   extra = TRUE,
#   cache = NULL,
#   latest = NULL, #=1 
#   language = "en"
# )
# saveRDS(wdi_in, "filesScale/wdi_in.rds")

wdi_in <- readRDS("filesScale/wdi_in.rds")


wdi_sel <- wdi_in %>%
  filter(region!="Aggregates")

  
p3 <- wdi_sel %>%
  mutate(sample='value', GDP=NY.GDP.MKTP.CD/1000000000, Population=SP.POP.TOTL/1000) %>%
  union(wdi_sel %>%
          mutate(sample='log', GDP=log10(NY.GDP.MKTP.CD/1000000000), Population=log10(SP.POP.TOTL/1000))) %>% 
  mutate(sample=factor(sample, levels=c("value", "log"))) %>%
  ggplot(aes(y=GDP, x=Population, text= country, color=region)) + #
  geom_point() +
  scale_colour_viridis_d(option = "turbo") +
  labs(y='GDP in billion USD', x='Population in thousand', title="Figure 2: Association between GDP and population size") +
    theme_bw() +
    theme(axis.text.y = element_text(angle=90)) +
  #, axis.title.x = element_text(margin=margin(t=60)), 
  #    axis.title.y = element_text(margin=margin(r=60))) +
  facet_wrap(~sample, scales = "free") 

ggplotly(p3)

```

A visual inspection of data (Figure 2) shows as expected: GDP increases as population grows. However, whereas a lot of variability and extreme values are found in the left plot where total values are shown, the double-logarithmic scale again reveals a linear relationship of the 2 variables.

```{r }
lm(log(NY.GDP.MKTP.CD) ~ log(SP.POP.TOTL), data=wdi_sel) %>%
  stargazer(type="text", title="Table 3: Association population size & GDP", align=TRUE, ci=TRUE)
```

The scaling exponent using country level data suggests sublinear scaling of about 0.8, ie. if the population doubles, GDP grows by 80 percent, on average. However, although interesting, it would again be helpful to understand the underlying mechanism, especially from a developmental perspective. Should the association of income and population always be analysed on a double-log scale? For example, the right hand side of Figure 2 is clearly helpful if one is interested in identifying single data points or if linear modeling is needed. But this representation also decreases variability and can potentially hide some important characteristics of the untransformed data. 

We have seen that data from a lot of different domains can be described by power laws, ie. associations become linear after applying the log to the variables of interest. This holds true for living organisms, cities, other institutions and even global structures such as countries.

---
nocite: |
  @p1, @p2, @p3, @p4
---


## References